Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-25 manuscript.rmd
1. Edited discussion

# Results
## Soil water dynamics by depth

The 100 ELM-FATES simulations with the chosen best-fit parameter sets (Table \@ref(tab:param)) successfully captured the dynamics of soil moisture by depth (Fig. \@ref(fig:hydro-calib) A & Fig. \@ref(fig:hydro-calib)B), including the observations from the horizontal TDR probes that were not included in the calibration. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)C) and in evapotranspiration (Fig. \@ref(fig:hydro-calib)D), slightly underestimating peak discharge in wet years and slightly overestimating peak ET. The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).

@@

\
\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought. The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity (also see, @Goulden.2019). In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa). The difference in the results shown here are likely attributable to site differences in the precipitation regime. In the dry forest site in CT2018 (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). The kind of droughts that buffered the deep rooted species in our study were repeated annual droughts that extended the dry season into the wet season. The variation in mortality explained by ERD in our study is upto 40%, and other causes may explain observed mortality [@McDowell.2018].

@@

@@

\
\   Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes. Future studies could better constrain the ERD model with higher resolution growth data such as those from dendrometer bands. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to reversible stem dehydration (xylem, phloem and bark), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015]. Use of dendrometer band data would thus entail accounting for reversible stem dehydration [@Zweifel:2018], but may also be an avenue to account for stem water storage and temporal dynamics of ERD, a limitation in our current approach. Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data for many species (Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and the timing of  leaf-gain backtracked from leaf lifetime, and omits inter-annual variation. Future studies could further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].

\
\   In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees. In sites where water does not limit growth, it may not be possible to parametarize ERD models from growth dynamics, and other avenues such as sap-flux or leaf water-potential dynamics could be explored [@Brum.2017]. While our ERD model empirically predicts growth via $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ at a specific depth, models that mechanistically account for $\mathrm{\Psi}$~soil~ and $\mathrm{\Psi}$~root~ over the entire rooting profile and then $\mathrm{\Psi}$~stem~ and $\mathrm{\Psi}$~leaf~, accounting for water content and embolised tissues, and downstream photosynthetic pathways and non-water limiting processes are likely to be more accurate in predicting growth [@Christoffersen:2016kd2d; @Sperry.1998]. Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g. retention curves, capacitance for roots, stems and leaves), adding a degree of parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g. see @Johnson:2018fad].

@@

>>>>

# Results
## Soil water dynamics by depth
\
The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig. \@ref(fig:hydro-calib)A, B), including the out of sample observations. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)C) and in evapotranspiration (Fig. \@ref(fig:hydro-calib)D), slightly underestimating peak discharge in wet years and slightly overestimating peak ET. The reduction in parameter range in the best-fit ensembles compared to the global ranges tested showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter (that regulates the depth of the rooting profile), soil hydraulic conductivity profile, especially at depth, and the adjustment factor that varies soil hydraulic conductivity to account for macroporosity and direct flow paths (Table \@ref(tab:param)).

@@

\
\   Our result--of lower mortality in deep rooted trees--contrasts with the finding of @Chitra-Tarak.2018 (hereafter CT2018) in which species identified by their inverse ERD model as effectively deep rooted died *more* in a rare, prolonged drought. The hydrological model in that study revealed that the multi-year drought exhausted the deep soil and even bedrock water availablity (also see, @Goulden.2019 and @Ivanov.2012). In contrast, our hydrological modeling at BCI finds that during the study period, $\mathrm{\Psi}$~soil~ below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and droughts, that is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa). The difference in the results shown here are likely attributable to site differences in the precipitation regime. In the dry forest site in CT2018 (Mudumalai National Park, India) precipitation is low: 1095 $\pm$ 296 mm (mean $\pm$ SD) compared to `r round(rain.man.stats$rain)` $\pm$ `r round(rain.man.stats$rain.sd)` mm (`r min(clim$Year)`-`r max(clim$Year)`) at BCI, and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (Penman-Monteith PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (Penman-Monteith PET = `r round(mean(clim$pet.PM))` $\pm$ `r round(sd(clim$pet.PM))` mm;`r min(clim$Year)`-`r max(clim$Year)`; this study, @Guo.2016), and thus at BCI deep soil layers are effectively recharged annually (Fig. S\@ref(fig:swp-all-depths); @Chitra-Tarak.2018). The kind of droughts that buffered the deep rooted species in our study were repeated annual droughts that extended the dry season into the wet season. The variation in mortality explained by ERD in our study is upto 40%, and other causes may explain observed mortality [@McDowell.2018].

@@

\
\   As our ERD model assumes diurnal maximum $\mathrm{\Psi}$~leaf~ as equivalent to $\mathrm{\Psi}$~soil~, ignoring the buffering effect of stem water storage capacitance, this approach may overestimate ERD in species where stem water storage capacitance is important as a drought-avoidance strategy, that is, the group of deciduous species [Borchert.2005jma].

@@

\
\   Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes. Future studies could better constrain the ERD model with higher resolution growth data such as those from dendrometer bands. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to reversible stem dehydration (xylem, phloem and bark), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015]. Use of dendrometer band data would thus entail accounting for reversible stem dehydration [@Zweifel:2018], but may also be an avenue to account for stem water storage and temporal dynamics of ERD, a limitation in our current approach. Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data for many species (Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and the timing of  leaf-gain backtracked from leaf lifetime, and omits inter-annual variation. Future studies could further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019]. This would also open the possibility of testing, whether deciduous species are more shallow rooted than evergreen species as generally observed [@Fan:2017].

\
\   In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees. In sites where water does not limit growth, it may not be possible to parametarize ERD models from growth dynamics, and other avenues such as sap-flux or leaf water-potential dynamics could be explored [@Brum.2017].

\
\   While our ERD model empirically predicts growth via $\mathrm{\Psi}$~leaf~ from $\mathrm{\Psi}$~soil~ at a specific depth, models that mechanistically account for $\mathrm{\Psi}$ through soil, the entire rooting profile, as well as, $\mathrm{\Psi}$~stem~ and $\mathrm{\Psi}$~leaf~, accounting for water content and accumulation of embolised tissues, and then downstream photosynthetic pathways and non-water limiting processes influencing growth are likely to be more accurate [@Christoffersen:2016kd2d; @Sperry.1998]. Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g. retention curves and capacitance for roots, stems and leaves), adding a degree of parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g. see @Johnson:2018fad].

@@

\
\ Finally, rooting depths may be parametarized within DGVMs such as ELM-FATES using plant hydrodynamic schemes [@Christoffersen:2016kd2d; @Koven.2020], simultaneously modeling plant water-availability with plant water-extraction. In general, including differences in tree rooting depths in ELM-FATES would improve understanding of plant competition for water (and other resources) and more accurately represent plant's drought strategies, leading to better predictions of existing and future forest function and composition.
